pdp: An R Package for Creating Partial Dependence Plots

### 2016-09-19 manuscript/greenwell.tex
minor edits

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(partial)
partial(boston.rf, pred.var = "lstat", plot = TRUE)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

@@

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region can be dangerous. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
\end{example}

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, "rm", plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than for loops. This makes it possible to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

@@

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.

@@

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
          chull = TRUE, train = subset(boston, select = -cmedv))
}
\end{example}

@@

PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors. In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches availble in other R packages as well. Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples. In terms of future development, \pkg{pdp} can be expanded in a number of ways. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response. It would also be nice to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.

>>>>

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
\begin{example}
library(pdp)
partial(boston.rf, pred.var = "lstat", plot = TRUE)
\end{example}
The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.

@@

In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted. Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised. The right display in Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
\end{example}

@@

Constructing PDPs can be quite computationally expensive. Several strategies are available to ease the computational burden in larger problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}. Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument. To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, "rm", plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

@@

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops. This makes it easy to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].

@@

It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}). Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.

@@

To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}. The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together. The results are displayed in Figure~\ref{fig:boston_xgb}. \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
\begin{example}
pred.list <- list("lstat", "rm", c("lstat", "rm"))
for (x in pred.list) {
  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
          chull = TRUE, train = subset(boston, select = -cmedv))
}
\end{example}

@@

PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors. In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package. We also briefly discussed related approaches available in other R packages. Suggestions to avoid extrapolation and high exececution times were discussed and demonstrated via examples.

In terms of future development, \pkg{pdp} can be expanded in a number of ways. For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response. It would also be worthwhile to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
